\documentclass[11pt]{article}

\usepackage[margin=1in]{geometry}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage{xcolor}
\usepackage{hyperref}
\usepackage{longtable}
\usepackage{tabularx}
\usepackage{array}
\usepackage{amsmath}
\usepackage{enumitem}

\hypersetup{
  colorlinks=true,
  linkcolor=blue,
  urlcolor=blue,
  pdftitle={shroomio Software Specification},
  pdfauthor={shroomio Authors}
}

\setlist[itemize]{leftmargin=1.5em}
\setlist[enumerate]{leftmargin=1.75em}
\setlength{\parskip}{0.5em}
\setlength{\parindent}{0pt}

\title{shroomio Software Specification\\\large Formal Product Requirements, Architecture, and Engineering Baseline}
\author{shroomio Authors}
\date{June 11, 2026}

\begin{document}

\maketitle

\tableofcontents

%---------------------------------------------------------------------
\section{Document Purpose}

This document is the formal software specification for \textbf{shroomio}, a real-time multiplayer fungal arena game written in C. It replaces the earlier informal \emph{Product Requirements and Technical Baseline} document and constitutes the single authoritative reference for all product requirements, technical architecture, protocol design, balancing parameters, and engineering roadmap.

This specification draws from three sources:
\begin{itemize}
  \item the actual implementation as it exists in the repository today,
  \item the intended product vision described in the earlier baseline document,
  \item a formal engineering review of the current codebase against that vision.
\end{itemize}

Where a requirement is already satisfied by the codebase it is annotated with the relevant source file. Where a requirement is not yet implemented it is marked as a scheduled milestone.

%---------------------------------------------------------------------
\section{Product Overview}

\subsection{Product Identity}

\textbf{shroomio} is a real-time multiplayer arena game in which players control fungal organisms, collect spores to grow, consume smaller organisms, avoid larger threats, and compete on a mass-based leaderboard. The game is designed to be playable both locally (offline against bots) and online (against human opponents via a headless dedicated server).

\subsection{Core Loop}

The intended player loop is:
\begin{enumerate}
  \item enter the arena at low-to-mid mass,
  \item collect safe spores in the outer ring,
  \item rotate toward mid and center zones to improve growth rate,
  \item watch for favorable fights against smaller targets,
  \item avoid larger predators as increased mass lowers mobility,
  \item climb the leaderboard,
  \item respawn quickly and re-engage after defeat.
\end{enumerate}

\subsection{Design Pillars}

\begin{itemize}
  \item \textbf{Readable competition}: player state and arena risk must be legible at a glance.
  \item \textbf{Fast re-entry}: defeat resets the player quickly without removing them from the session.
  \item \textbf{Authoritative multiplayer}: server state is the source of truth for movement and collisions.
  \item \textbf{Low content overhead}: the game remains fun with simple shapes, simple rules, and lightweight assets.
  \item \textbf{Incremental extensibility}: simulation, client rendering, and server networking stay separated so the prototype can evolve safely.
\end{itemize}

%---------------------------------------------------------------------
\section{Architecture}

\subsection{Module Map}

The repository is organized into four top-level modules, each with a defined responsibility and dependency boundary.

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.18\textwidth}|>{\raggedright\arraybackslash}p{0.32\textwidth}|>{\raggedright\arraybackslash}p{0.38\textwidth}|}
\hline
\textbf{Module} & \textbf{Source Path} & \textbf{Responsibility} \\
\hline
Client / Main & \texttt{src/client/main.c} & Application lifecycle, window creation, main loop, state machine dispatch \\
\hline
Client / Screens & \texttt{src/client/screens/} & Screen state machine (menu, lobby, game, settings, etc.) \\
\hline
Client / Game & \texttt{src/client/game.c}, \texttt{game.h} & Game state initialization, update tick, 2D rendering, camera, input \\
\hline
Client / Net & \texttt{src/client/net.c}, \texttt{net.h} & ENet lifecycle, handshake, input sending, snapshot reception \\
\hline
Client / Chat & \texttt{src/client/chat.c}, \texttt{chat.h} & Text chat UI, message history, input handling \\
\hline
Client / Voice & \texttt{src/client/voice.c}, \texttt{voice.h} & Microphone capture, Opus encoding/decoding, audio playback \\
\hline
Client / Auth & \texttt{src/client/auth.c}, \texttt{auth.h} & Authentication UI, token storage, login/logout \\
\hline
Server / Main & \texttt{src/server/main.c} & Server lifecycle, ENet host, session management, tick loop \\
\hline
Server / Auth & \texttt{src/server/auth.c}, \texttt{auth.h} & Authentication validation, token management, Discord OAuth \\
\hline
Server / Chat & \texttt{src/server/chat.c}, \texttt{chat.h} & Chat message routing, rate limiting, moderation \\
\hline
Server / Voice & \texttt{src/server/voice.c}, \texttt{voice.h} & Voice packet routing, mixing, bandwidth management \\
\hline
Server / Logger & \texttt{src/server/logger.c}, \texttt{logger.h} & Structured logging with levels, timestamps, colored output \\
\hline
Server / Master & \texttt{src/server/master.c}, \texttt{master.h} & Master server registration, heartbeat, server discovery \\
\hline
Shared / Config & \texttt{src/shared/config.h} & Compile-time tuning constants \\
\hline
Shared / World & \texttt{src/shared/world.h} & Data types: world state, player state, spore state, zones \\
\hline
Shared / Sim & \texttt{src/shared/sim.c}, \texttt{sim.h} & Deterministic simulation: movement, collection, consume, bot AI \\
\hline
Shared / Protocol & \texttt{src/shared/protocol.h} & Binary packet formats for all client-server messages \\
\hline
Shared / Vector & \texttt{src/shared/vec2.h} & Inline 2D vector math \\
\hline
\end{longtable}

\subsection{Dependency Graph}

\begin{verbatim}
client/main.c
  -> client/screens/* -> client/game.h -> client/net.h -> shared/protocol.h
                    -> client/chat.h                     -> shared/vec2.h
                    -> client/voice.h -> shared/world.h -> shared/config.h
                    -> client/auth.h                   -> raylib, opus, enet

server/main.c
  -> server/auth.h    -> server/logger.h
  -> server/chat.h    -> shared/sim.h    -> shared/world.h -> shared/config.h
  -> server/voice.h   -> shared/protocol.h                -> shared/vec2.h
  -> server/master.h  -> enet, sqlite3, opus

shared/sim.c
  -> shared/sim.h
  -> shared/world.h
\end{verbatim}

\noindent\textbf{Rule}: client code may depend on shared code but never on server code. Server code may depend on shared code but never on client code. Shared code has no dependency on either client or server, and no dependency on \texttt{raylib}, \texttt{enet}, or \texttt{opus}.

\subsection{Build System}

The project uses a single GNU Makefile at the repository root.

\begin{itemize}
  \item \texttt{make linux} --- builds the Linux raylib client binary (\texttt{dist/shroomio}).
  \item \texttt{make server} --- builds the Linux headless server binary (\texttt{dist/shroomio-server}).
  \item \texttt{make windows} --- cross-compiles the Windows client binary (\texttt{dist/shroomio.exe}) using mingw-w64.
  \item \texttt{make run} --- builds and launches the Linux client.
  \item \texttt{make run-server} --- builds and launches the headless server.
  \item \texttt{make docker-server} --- builds the server Docker image.
  \item \texttt{make docker-run-server} --- builds and runs the server Docker container.
\end{itemize}

The Makefile downloads raylib source (version 5.5) on demand, compiles \texttt{libraylib.a} statically for each target, and bundles ENet (vendored under \texttt{vendor/enet/}) for the server target. The server links against SQLite (\texttt{-lsqlite3}) for persistent storage.

\noindent\textbf{Supported platforms}:
\begin{itemize}
  \item Linux client (gcc/clang, raylib static, system GL/X11 libs dynamic)
  \item Linux headless server (gcc/clang, no graphics libraries)
  \item Windows client cross-build from Linux (mingw-w64, fully static exe)
\end{itemize}

%---------------------------------------------------------------------
\section{Data Structures}

\subsection{World State}

\begin{verbatim}
ShroomWorldState {
    uint64_t             tick;
    float                width, height;          // 6000 x 6000
    ShroomEntityId       next_entity_id;
    size_t               player_count;           // <= SHROOM_MAX_PLAYERS (256)
    size_t               spore_count;            // <= SHROOM_MAX_SPORES (4096)
    ShroomPlayerState    players[256];
    ShroomSporeState     spores[4096];
}
\end{verbatim}

\subsection{Player State}

\begin{verbatim}
ShroomPlayerState {
    ShroomPlayerId  player_id;        // unique player identifier
    ShroomEntityId  entity_id;        // unique entity identifier
    ShroomVec2      position;         // world-space position
    ShroomVec2      input_direction;  // normalized or zero
    float           mass;             // current mass
    float           radius;           // derived from mass
    bool            alive;
    bool            is_bot;
}
\end{verbatim}

\subsection{Spore State}

\begin{verbatim}
ShroomSporeState {
    ShroomEntityId  entity_id;
    ShroomVec2      position;
    uint16_t        value;            // mass granted on collection (8)
    bool            active;
}
\end{verbatim}

\subsection{Zone Enumeration}

\begin{verbatim}
ShroomZone {
    SHROOM_ZONE_OUTER  = 0,
    SHROOM_ZONE_MID    = 1,
    SHROOM_ZONE_CENTER = 2,
}
\end{verbatim}

Zones are concentric circles centered on the world midpoint. Center radius is 900, mid radius is 2000, outer is everything beyond mid up to the world border.

%---------------------------------------------------------------------
\section{Database Schema}

The server uses SQLite for persistent storage of player statistics, session history, and match events. The database schema is versioned and located in \texttt{database/migrations/}.

\subsection{Schema Overview}

The data model consists of seven core tables and two leaderboard views:

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.25\textwidth}|>{\raggedright\arraybackslash}p{0.65\textwidth}|}
\hline
\textbf{Table} & \textbf{Purpose} \\
\hline
\texttt{players} & Tracks all known players by UUID and display name \\
\hline
\texttt{users} & Authentication accounts (username, password hash, Discord ID) \\
\hline
\texttt{sessions} & Records individual game sessions with timing and participant counts \\
\hline
\texttt{session\_participants} & Links players to sessions with join/leave times and final stats \\
\hline
\texttt{player\_stats} & Cumulative statistics per player (kills, deaths, mass, etc.) \\
\hline
\texttt{match\_events} & Significant in-game events (spawns, deaths, consumptions) \\
\hline
\texttt{auth\_tokens} & Active authentication tokens for logged-in users \\
\hline
\end{longtable}

\subsection{Players Table}

\begin{verbatim}
CREATE TABLE players (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    player_uuid TEXT NOT NULL UNIQUE,
    display_name TEXT NOT NULL,
    created_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    last_seen_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    total_sessions INTEGER NOT NULL DEFAULT 0,
    total_play_time_seconds INTEGER NOT NULL DEFAULT 0
);
\end{verbatim}

Each player is uniquely identified by a UUID. The \texttt{display\_name} is the player-chosen name sent in the Hello packet. The server updates \texttt{last\_seen\_at} on each session participation.

\subsection{Users Table (Authentication)}

\begin{verbatim}
CREATE TABLE users (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    player_id INTEGER NOT NULL UNIQUE,
    username TEXT NOT NULL UNIQUE,
    email TEXT,
    password_hash TEXT,
    discord_id TEXT UNIQUE,
    auth_method TEXT NOT NULL CHECK (auth_method IN ('password', 'discord', 'anonymous')),
    created_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    last_login_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    FOREIGN KEY (player_id) REFERENCES players(id) ON DELETE CASCADE
);
\end{verbatim}

Each user account is linked to a player record. The \texttt{auth\_method} indicates how the account was created. Password-based accounts have a \texttt{password\_hash} (bcrypt or argon2). Discord accounts have a \texttt{discord\_id}.

\subsection{Auth Tokens Table}

\begin{verbatim}
CREATE TABLE auth_tokens (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    user_id INTEGER NOT NULL,
    token TEXT NOT NULL UNIQUE,
    token_type TEXT NOT NULL DEFAULT 'jwt',
    expires_at TEXT NOT NULL,
    created_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    revoked_at TEXT,
    FOREIGN KEY (user_id) REFERENCES users(id) ON DELETE CASCADE
);
\end{verbatim}

Active authentication tokens are stored server-side for validation and revocation. Tokens expire after 24 hours by default.

\subsection{Sessions Table}

\begin{verbatim}
CREATE TABLE sessions (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    session_uuid TEXT NOT NULL UNIQUE,
    started_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    ended_at TEXT,
    duration_seconds INTEGER,
    player_count INTEGER NOT NULL DEFAULT 0,
    bot_count INTEGER NOT NULL DEFAULT 0,
    status TEXT NOT NULL CHECK (status IN ('active', 'completed', 'aborted')) DEFAULT 'active'
);
\end{verbatim}

Sessions are created when the server starts a new game instance. The \texttt{status} field tracks whether the session is currently running (\texttt{active}), finished normally (\texttt{completed}), or terminated early (\texttt{aborted}).

\subsection{Session Participants Table}

\begin{verbatim}
CREATE TABLE session_participants (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    session_id INTEGER NOT NULL,
    player_id INTEGER NOT NULL,
    joined_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    left_at TEXT,
    final_rank INTEGER,
    final_mass REAL,
    FOREIGN KEY (session_id) REFERENCES sessions(id) ON DELETE CASCADE,
    FOREIGN KEY (player_id) REFERENCES players(id) ON DELETE CASCADE,
    UNIQUE(session_id, player_id)
);
\end{verbatim}

This table records which players participated in each session. The \texttt{final\_rank} and \texttt{final\_mass} are populated when the session ends.

\subsection{Player Statistics Table}

\begin{verbatim}
CREATE TABLE player_stats (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    player_id INTEGER NOT NULL UNIQUE,
    total_games_played INTEGER NOT NULL DEFAULT 0,
    total_kills INTEGER NOT NULL DEFAULT 0,
    total_deaths INTEGER NOT NULL DEFAULT 0,
    total_mass_consumed REAL NOT NULL DEFAULT 0.0,
    total_mass_lost REAL NOT NULL DEFAULT 0.0,
    total_distance_traveled REAL NOT NULL DEFAULT 0.0,
    total_spores_consumed INTEGER NOT NULL DEFAULT 0,
    total_players_consumed INTEGER NOT NULL DEFAULT 0,
    highest_mass_achieved REAL NOT NULL DEFAULT 0.0,
    highest_rank_achieved INTEGER NOT NULL DEFAULT 999,
    longest_survival_seconds REAL NOT NULL DEFAULT 0.0,
    updated_at TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    FOREIGN KEY (player_id) REFERENCES players(id) ON DELETE CASCADE
);
\end{verbatim}

This table maintains cumulative statistics for each player. It is updated at the end of each session.

\subsection{Match Events Table}

\begin{verbatim}
CREATE TABLE match_events (
    id INTEGER PRIMARY KEY AUTOINCREMENT,
    session_id INTEGER NOT NULL,
    event_type TEXT NOT NULL CHECK (event_type IN (
        'player_spawn',
        'player_death',
        'player_consume_player',
        'player_consume_spore',
        'player_reach_mass',
        'game_tick'
    )),
    event_timestamp TEXT NOT NULL DEFAULT (strftime('%Y-%m-%dT%H:%M:%SZ', 'now')),
    tick_number INTEGER,
    actor_player_id INTEGER,
    target_player_id INTEGER,
    mass_value REAL,
    position_x REAL,
    position_y REAL,
    metadata TEXT,
    FOREIGN KEY (session_id) REFERENCES sessions(id) ON DELETE CASCADE,
    FOREIGN KEY (actor_player_id) REFERENCES players(id) ON DELETE SET NULL,
    FOREIGN KEY (target_player_id) REFERENCES players(id) ON DELETE SET NULL
);
\end{verbatim}

This table logs significant events during gameplay for analytics and replay purposes. The \texttt{metadata} field is a JSON string for extensible event-specific data.

\subsection{Leaderboard Views}

Two views provide convenient access to top players:

\begin{verbatim}
CREATE VIEW leaderboard_by_mass AS
SELECT 
    p.display_name,
    ps.highest_mass_achieved,
    ps.total_games_played,
    ps.total_kills,
    p.last_seen_at
FROM players p
JOIN player_stats ps ON p.id = ps.player_id
WHERE ps.highest_mass_achieved > 0
ORDER BY ps.highest_mass_achieved DESC;

CREATE VIEW leaderboard_by_kills AS
SELECT 
    p.display_name,
    ps.total_kills,
    ps.total_players_consumed,
    ps.total_games_played,
    p.last_seen_at
FROM players p
JOIN player_stats ps ON p.id = ps.player_id
WHERE ps.total_kills > 0
ORDER BY ps.total_kills DESC;
\end{verbatim}

\subsection{Database Location and Initialization}

The SQLite database file is stored at \texttt{data/shroomio.db} by default. The server creates the database and applies migrations on first startup if the file does not exist. Migration files are located in \texttt{database/migrations/} and are applied in lexicographic order.

%---------------------------------------------------------------------
\section{Protocol Specification}

\subsection{Transport}

All multiplayer communication uses ENet over UDP on port 7777. ENet provides multiple channels with different quality-of-service guarantees.

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.18\textwidth}|>{\raggedright\arraybackslash}p{0.12\textwidth}|>{\raggedright\arraybackslash}p{0.16\textwidth}|>{\raggedright\arraybackslash}p{0.38\textwidth}|}
\hline
\textbf{Channel} & \textbf{ID} & \textbf{QoS} & \textbf{Purpose} \\
\hline
Control & 0 & Reliable, ordered & Hello, Welcome, Ping, Pong, Auth, Chat \\
\hline
Input & 1 & Unreliable, sequenced & Player directional input \\
\hline
Snapshot & 2 & Unreliable, sequenced & Server-to-client world state updates \\
\hline
Chat & 3 & Reliable, ordered & Text chat messages \\
\hline
Voice & 4 & Unreliable, ordered & Voice chat packets (Opus encoded) \\
\hline
\end{longtable}

\subsection{Common Header}

Every packet begins with a 4-byte header:

\begin{verbatim}
ShroomPacketHeader {
    uint8_t  type;       // ShroomPacketType enum
    uint8_t  reserved;
    uint16_t size;       // total packet size in bytes
}
\end{verbatim}

\subsection{Packet Types}

\subsubsection{Hello (client to server, type = 1)}

\begin{verbatim}
ShroomHelloPacket {
    ShroomPacketHeader header;
    uint32_t protocol_version;    // SHROOM_PROTOCOL_VERSION (1)
    char     name[32];            // player-chosen display name
}
\end{verbatim}

Sent immediately after ENet connection. If \texttt{protocol\_version} does not match, the server disconnects the peer. If the server is full, the peer is disconnected.

\subsubsection{Welcome (server to client, type = 2)}

\begin{verbatim}
ShroomWelcomePacket {
    ShroomPacketHeader header;
    uint32_t protocol_version;
    uint32_t player_id;
    uint32_t entity_id;
    uint16_t server_tick_rate;    // 30
    uint16_t snapshot_rate;       // 15
    float    world_width;         // 6000.0
    float    world_height;        // 6000.0
}
\end{verbatim}

Sent in response to a valid Hello. After receiving Welcome, the client switches from local simulation to applying server snapshots.

\subsubsection{Input (client to server, type = 3)}

\begin{verbatim}
ShroomInputPacket {
    ShroomPacketHeader header;
    uint32_t sequence;            // monotonic, client-side
    float    direction_x;         // normalized input x
    float    direction_y;         // normalized input y
}
\end{verbatim}

Sent per tick at the server tick rate (30~Hz). The server clamps and normalizes the direction vector before applying it.

\subsubsection{Snapshot (server to client, type = 4)}

\begin{verbatim}
ShroomSnapshotPacket {
    ShroomPacketHeader              header;
    uint64_t                        tick;
    uint32_t                        last_processed_input_sequence;
    uint32_t                        player_id;        // this client's player
    uint32_t                        entity_id;
    uint16_t                        player_count;
    uint16_t                        reserved;
    ShroomSnapshotPlayerState       players[32];
}

ShroomSnapshotPlayerState {
    uint32_t player_id;
    uint32_t entity_id;
    float    position_x;
    float    position_y;
    float    mass;
    float    radius;
    uint8_t  alive;
    uint8_t  is_bot;
    uint16_t reserved;
}
\end{verbatim}

Broadcast at the snapshot rate (15~Hz). Contains up to 32 player entries. Clients overwrite their local player array from this payload.

\noindent\textbf{Known limitation}: spores are not included in snapshot packets. Online players do not see or interact with spores through the network layer. Spore replication is milestone 1 in the engineering roadmap (\S\ref{sec:roadmap}).

\subsubsection{Ping / Pong (type = 5, 6)}

\begin{verbatim}
ShroomPingPacket  { ShroomPacketHeader; uint32_t nonce; }
ShroomPongPacket  { ShroomPacketHeader; uint32_t nonce; }
\end{verbatim}

Used for latency measurement. The server echoes Pong with the same nonce. Not yet wired to a visible RTT display on the client.

\subsection{Snapshot Capacity Constraint}

The simulation supports up to 128 players (\texttt{SHROOM\_MAX\_PLAYERS}), but the snapshot packet is hard-coded to 32 player entries (\texttt{SHROOM\_MAX\_SNAPSHOT\_PLAYERS}). This means that in a full online match, only the first 32 alive players will be visible to any given client. This is a known capacity gap scheduled for remediation in milestone 2 (\S\ref{sec:roadmap}).

%---------------------------------------------------------------------
\section{Simulation Rules}

\subsection{World}

\begin{itemize}
  \item World dimensions: $6000 \times 6000$ units (square).
  \item Entity positions are clamped to the rectangle $[radius,\; width - radius] \times [radius,\; height - radius]$.
  \item Three concentric zones centered at $(3000, 3000)$: center~($r \le 900$), mid~($900 < r \le 2000$), outer~($r > 2000$).
\end{itemize}

\subsection{Movement}

\begin{itemize}
  \item Each tick, each alive player is moved by $\text{velocity} = \text{input\_direction} \times \text{speed} \times \Delta t$.
  \item Input direction is a normalized 2D vector (zero allowed for no movement).
  \item Speed is derived from mass:
    \begin{verbatim}
speed = SHROOM_MAX_PLAYER_SPEED / (1.0 + (mass * 0.35 / 100.0))
clamp(speed, SHROOM_MIN_PLAYER_SPEED, SHROOM_MAX_PLAYER_SPEED)
    \end{verbatim}
  \item Minimum speed: 120. Maximum speed: 260. (Both units/sec.)
  \item Bots compute their own input direction each tick (see \S\ref{sec:bot-ai}).
\end{itemize}

\subsection{Spore Collection}

\begin{itemize}
  \item The world maintains a target of 900 active spores.
  \item Each spore has a value of 8 mass.
  \item Collection radius is $\text{player\_radius} + 6.0$ units.
  \item When a player overlaps a spore, the player gains the spore's mass value and the spore is immediately relocated to a new random position (biased toward center and mid zones).
\end{itemize}

\noindent\textbf{Spore spawn bias} (per \texttt{ShroomSpawnOrResetSpore}):
\begin{itemize}
  \item 45\% chance: placed within the center zone.
  \item 35\% chance: placed within the mid zone (beyond center but within mid radius).
  \item 20\% chance: placed anywhere in the world (outer).
\end{itemize}

\subsection{Consumption}

\begin{itemize}
  \item A player may consume another player if:
    \begin{enumerate}
      \item both are alive,
      \item the attacker's mass is at least $1.15 \times$ the target's mass,
      \item the attacker's center overlaps the target's radius scaled by $0.88$.
    \end{enumerate}
  \item On consumption: the attacker gains $0.70 \times$ the victim's mass; the victim respawns immediately at default mass (100) in a safe outer-zone location.
  \item Multiple simultaneous consumes are resolved in a single pass: a victim is consumed at most once per tick.
\end{itemize}

\subsection{Spawning and Respawning}

\begin{itemize}
  \item New players and respawning players are placed in the outer zone, at least 340 units away from any alive player.
  \item Up to 32 random candidate positions are tried before falling back to a random position without safety checks.
  \item Default spawn mass is 100.
  \item Radius is recalculated from mass: $\text{radius} = 10.0 + \text{mass} \times 0.14$.
\end{itemize}

\subsection{Bot AI}
\label{sec:bot-ai}

Bots follow a priority-ordered behavior tree evaluated each tick:

\begin{enumerate}
  \item \textbf{Flee}: if a larger player (mass $>$ bot\_mass $\times 1.15$) is within 900 units, move directly away.
  \item \textbf{Chase}: if a smaller player (bot\_mass $>$ target\_mass $\times 1.15$) is within 700 units, move directly toward.
  \item \textbf{Forage}: move toward the nearest active spore.
  \item \textbf{Drift}: move toward the world center.
\end{enumerate}

\subsection{Tuning Parameters}

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.48\textwidth}|>{\raggedright\arraybackslash}p{0.20\textwidth}|>{\raggedright\arraybackslash}p{0.20\textwidth}|}
\hline
\textbf{Parameter} & \textbf{Value} & \textbf{Defined In} \\
\hline
World width / height & $6000 \times 6000$ & \texttt{config.h:7--8} \\
\hline
Max players (sim) & 256 & \texttt{config.h:4} \\
\hline
Max spores & 4096 & \texttt{config.h:5} \\
\hline
Max snapshot players & 32 & \texttt{protocol.h:15} \\
\hline
Center zone radius & 900 & \texttt{config.h:10} \\
\hline
Mid zone radius & 2000 & \texttt{config.h:11} \\
\hline
Default player mass & 100 & \texttt{config.h:13} \\
\hline
Bot count & 18 & \texttt{config.h:14} \\
\hline
Min player speed & 120 & \texttt{config.h:15} \\
\hline
Max player speed & 260 & \texttt{config.h:16} \\
\hline
Speed mass scale & 0.35 & \texttt{config.h:17} \\
\hline
Spore target count & 900 & \texttt{config.h:19} \\
\hline
Spore value & 8 & \texttt{config.h:20} \\
\hline
Consume mass advantage & 1.15 & \texttt{config.h:22} \\
\hline
Consume mass gain factor & 0.70 & \texttt{config.h:23} \\
\hline
Spawn safe distance & 340 & \texttt{config.h:24} \\
\hline
Server tick rate & 30 Hz & \texttt{config.h:26} \\
\hline
Snapshot rate & 15 Hz & \texttt{protocol.h:14} \\
\hline
Server port & 7777 / UDP & \texttt{protocol.h:7} \\
\hline
Protocol version & 1 & \texttt{protocol.h:6} \\
\hline
Max player name length & 32 & \texttt{protocol.h:8} \\
\hline
Latency goal & < 100ms & \texttt{config.h} \\
\hline
Latency warning & 150ms & \texttt{config.h} \\
\hline
Latency unplayable & > 200ms & \texttt{config.h} \\
\hline
Voice bitrate & 32--64kbps & \texttt{config.h} \\
\hline
Voice sample rate & 48kHz & \texttt{config.h} \\
\hline
Voice frame size & 20ms & \texttt{config.h} \\
\hline
Chat max length & 200 chars & \texttt{protocol.h} \\
\hline
Chat rate limit & 5 msgs/10 sec & \texttt{config.h} \\
\hline
\end{longtable}

%---------------------------------------------------------------------
\section{Client Specification}

\subsection{Rendering}

\begin{itemize}
  \item Uses \texttt{raylib} 5.5 with a 2D camera (\texttt{Camera2D}) centered on the local player.
  \item Target window: 1280$\times$720 at 60 FPS.
  \item Arena zones rendered as colored rectangles/circles with alpha blending.
  \item Player entities rendered as filled circles, color-coded: human players in green (\texttt{\{126,217,87\}}), bots in preset palette colors rotated by player ID.
  \item Local player highlighted with an additional white outline ring.
  \item Spores rendered as small gold circles (radius 4).
  \item Grid drawn every 80 units for spatial orientation.
  \item World border drawn as a rectangle outline.
\end{itemize}

\subsection{HUD}

The client renders two fixed HUD panels (top-left and top-right) regardless of camera position:

\begin{itemize}
  \item \textbf{Top-left}:
    \begin{itemize}
      \item Title.
      \item Movement hint (``Move toward the mouse\ldots'').
      \item Mass, radius, current speed.
      \item Zone label, player count, spore count, tick.
      \item Server connection status and last snapshot tick (green when connected, orange otherwise).
    \end{itemize}
  \item \textbf{Top-right}: leaderboard, top 6 entries, highlighting the local player with ``You'' label and white text.
\end{itemize}

\subsection{Input}

Primary movement: mouse-to-world vector (screen to world using \texttt{GetScreenToWorld2D}). Secondary: WASD and arrow keys add $+1.0$ or $-1.0$ to the accumulated direction vector. The resulting vector is normalized before being sent to the simulation or the server.

\subsection{Dual-Mode Operation}

\begin{itemize}
  \item \textbf{Offline mode}: the client runs the shared simulation (\texttt{ShroomWorldStep}) locally. No network connection required. Full spore economy and bot AI run client-side.
  \item \textbf{Online mode}: after a successful Welcome handshake, the client stops running \texttt{ShroomWorldStep} locally and instead applies snapshots from the server (\texttt{ApplyNetworkSnapshot}). Spore state is zeroed in this mode.
\end{itemize}

\subsection{Network Client Lifecycle}

\begin{verbatim}
1. ENet init
2. Create ENet client host
3. Resolve server address
4. Connect to server
5. Wait for ENet connect event -> send Hello
6. Wait for Welcome -> status = CONNECTED
7. Send Input packets each tick at 30 Hz
8. Receive Snapshot packets at ~15 Hz
9. On disconnect -> reset to local offline mode
\end{verbatim}

%---------------------------------------------------------------------
\section{Server Specification}

\subsection{Startup}

\begin{verbatim}
1. Initialize logger (INFO level, colored output)
2. Install SIGINT / SIGTERM handlers
3. enet_initialize()
4. Initialize database (create/open SQLite file, apply migrations)
5. ShroomWorldInit() -> spawn 18 bots
6. enet_host_create(&address, 256 clients, 5 channels)
7. Create new session record in database
8. Register with master server (if applicable)
9. Enter tick loop
\end{verbatim}

\subsection{Tick Loop}

\begin{itemize}
  \item Fixed-timestep loop at 30 Hz using \texttt{clock\_gettime(CLOCK\_MONOTONIC)} and \texttt{nanosleep}.
  \item Each tick:
    \begin{enumerate}
      \item Service all ENet events (connect, receive, disconnect).
      \item Call \texttt{ShroomWorldStep(world, 1/30)}.
      \item If tick is divisible by the snapshot interval (every 2nd tick), broadcast snapshots to all connected peers.
      \item Flush ENet.
      \item Sleep until next tick boundary.
    \end{enumerate}
\end{itemize}

\subsection{Session Management}

\begin{itemize}
  \item Each connected peer is assigned a \texttt{ServerSession} slot (max 256).
  \item On connect: the session is stored via \texttt{peer->data}.
  \item On Hello: session is activated, player spawned, Welcome sent.
  \item On Input: the input sequence is recorded and direction applied to the player.
  \item On disconnect: the player is zeroed (mass = 0, radius = 0, alive = false) and the session deactivated.
  \item Connection health tracked: ping, packet loss, RTT average.
  \item Rate limiting on rapid reconnect attempts.
\end{itemize}

\subsection{Snapshot Broadcast}

Snapshots are sent to every connected peer. The snapshot payload includes the first 32 alive players with positive mass. If more than 32 alive players exist, only the first 32 are included.

\subsection{Deployment}

The server is packaged as a Docker container:

\begin{verbatim}
FROM debian:bookworm-slim AS builder
# install build-essential, make
# COPY Makefile, src/, vendor/enet
# RUN make server

FROM debian:bookworm-slim
# COPY shroomio-server binary
# EXPOSE 7777/udp
# CMD ["/usr/local/bin/shroomio-server"]
\end{verbatim}

\noindent Run with:
\begin{verbatim}
make docker-server      # build image
make docker-run-server  # build and run
docker compose up --build server  # via compose
\end{verbatim}

%---------------------------------------------------------------------
\section{Authentication System}

\subsection{Authentication Methods}

The game supports three authentication methods, ordered by preference:

\begin{enumerate}
  \item \textbf{Username + Password} (encouraged): Traditional account-based authentication with persistent identity.
  \item \textbf{Discord OAuth} (encouraged): OAuth2 flow with Discord for seamless social login.
  \item \textbf{Anonymous} (available): Pick any username, no password required. Limited features.
\end{enumerate}

\subsection{Authentication Flow}

\begin{verbatim}
Client                          Server
  |                                |
  |-- AuthRequest (method, creds) ->|
  |                                |-- Validate credentials
  |                                |-- Create/find player record
  |<-- AuthResponse (token, info) -|
  |                                |
  |-- Hello (token, name) -------->|
  |                                |-- Validate token
  |<-- Welcome (player_id, ...) --|
\end{verbatim}

\subsection{Token Management}

Authenticated users receive a JWT (JSON Web Token) with:
\begin{itemize}
  \item Player UUID
  \item Username
  \item Authentication method
  \item Expiration timestamp (24 hours)
  \item Signature (HMAC-SHA256)
\end{itemize}

Tokens are stored client-side in \texttt{\textasciitilde/.config/shroomio/auth.json} with restrictive permissions (0600).

\subsection{Anonymous Mode}

Anonymous players:
\begin{itemize}
  \item Can join with any username (not reserved)
  \item No persistent stats or identity
  \item Marked as "Guest" in UI
  \item Cannot access ranked features (future)
\end{itemize}

%---------------------------------------------------------------------
\section{Voice Chat}

\subsection{Audio Pipeline}

\begin{enumerate}
  \item \textbf{Capture}: Record microphone input (platform-specific: ALSA on Linux, WASAPI on Windows)
  \item \textbf{Encode}: Compress with Opus codec (32--64kbps, 48kHz, 20ms frames)
  \item \textbf{Transmit}: Send encoded packets over UDP (Channel 4, unreliable)
  \item \textbf{Receive}: Decode incoming voice packets
  \item \textbf{Playback}: Mix and play through speakers
\end{enumerate}

\subsection{Opus Codec Configuration}

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.40\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|}
\hline
\textbf{Parameter} & \textbf{Value} \\
\hline
Sample rate & 48kHz \\
\hline
Bitrate & 32--64kbps (configurable) \\
\hline
Frame size & 20ms \\
\hline
Complexity & 5--8 (quality vs CPU) \\
\hline
Channels & Mono \\
\hline
Application & VOIP \\
\hline
\end{longtable}

\subsection{Voice Chat Modes}

\begin{itemize}
  \item \textbf{Push-to-talk (PTT)}: Hold key to transmit (default: V or B). Recommended for v1.
  \item \textbf{Always-on}: Continuous transmission with voice activity detection (VAD). Optional.
\end{itemize}

\subsection{Bandwidth Budget}

\begin{itemize}
  \item Per active speaker: 30--50kbps
  \item Receiving client (5--10 simultaneous speakers): 150--500kbps
  \item Total voice bandwidth per client: $\sim$200kbps typical
\end{itemize}

\subsection{Voice Features}

\begin{itemize}
  \item Mute/unmute self
  \item Mute individual players
  \item Volume control per speaker
  \item Voice activity indicator (who's talking)
  \item Push-to-talk key binding (configurable)
  \item Automatic gain control (AGC)
  \item Jitter buffer for incoming audio
\end{itemize}

%---------------------------------------------------------------------
\section{Text Chat}

\subsection{Chat Types}

\begin{itemize}
  \item \textbf{Global chat}: Messages visible to all players in lobby
  \item \textbf{Team chat}: Messages visible to team members only (future, if teams implemented)
\end{itemize}

\subsection{Chat Features}

\begin{itemize}
  \item Send/receive text messages
  \item Display chat history (last 50 messages)
  \item Show sender name and timestamp
  \item Message length limit: 200 characters
  \item Rate limiting: max 5 messages per 10 seconds per player
  \item Mute individual players
  \item Profanity filter (optional, configurable)
\end{itemize}

\subsection{Chat Protocol}

\begin{verbatim}
ShroomChatPacket {
    ShroomPacketHeader header;
    uint32_t           sender_id;
    uint64_t           timestamp;
    char               message[200];
}
\end{verbatim}

Sent on Channel 3 (reliable, ordered) to ensure no messages are lost.

%---------------------------------------------------------------------
\section{Server Discovery and Matchmaking}

\subsection{Master Server}

A lightweight HTTP service that tracks active game servers:

\begin{itemize}
  \item Game servers register on startup with: IP, port, name, max players, map
  \item Heartbeat every 30 seconds with updated stats (current player count)
  \item Deregister on graceful shutdown
  \item Clients query master server for server list
\end{itemize}

\subsection{Lobby Browser}

Client UI for browsing available servers:

\begin{itemize}
  \item Display: server name, player count (current/max), ping, map
  \item Sort by: name, player count, ping
  \item Filter by: player count range, map
  \item Refresh button
  \item Direct connect option (IP:port)
\end{itemize}

\subsection{Basic Matchmaking}

Automatic server assignment based on:

\begin{enumerate}
  \item \textbf{Primary}: Lowest ping/latency (best connection quality)
  \item \textbf{Secondary}: Lowest player count (balance load)
  \item \textbf{Tertiary}: Server has space (< 256 players)
\end{enumerate}

\textbf{Scoring algorithm}:
\begin{verbatim}
score = (0.7 * ping_ms) + (0.3 * player_count)
\end{verbatim}

Lower score = better choice. Servers with ping > 200ms are excluded.

%---------------------------------------------------------------------
\section{Performance Targets}

\subsection{Latency Budget}

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.30\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|}
\hline
\textbf{Threshold} & \textbf{RTT} & \textbf{Action} \\
\hline
Goal & < 100ms & Smooth gameplay \\
\hline
Warning & 150ms & Show "high latency" warning \\
\hline
Unplayable & > 200ms & Exclude from matchmaking, warn player \\
\hline
\end{longtable}

\subsection{Throughput Targets}

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.40\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|}
\hline
\textbf{Metric} & \textbf{Target} \\
\hline
Input rate (client $\rightarrow$ server) & 30 Hz (30 msgs/sec) \\
\hline
Snapshot rate (server $\rightarrow$ client) & 15--20 Hz \\
\hline
Total per client & 45--50 msgs/sec \\
\hline
Server total (256 players) & 11,500--12,800 msgs/sec \\
\hline
Voice per speaker & 30--50kbps \\
\hline
Total per client (with voice) & $\sim$200kbps \\
\hline
\end{longtable}

\subsection{Server Hardware Requirements}

Target specification for 256-player lobbies with voice chat:

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.30\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|}
\hline
\textbf{Component} & \textbf{Specification} \\
\hline
CPU & 4 cores (8 threads recommended) \\
\hline
RAM & 8GB minimum (16GB recommended) \\
\hline
Network & 1Gbps uplink \\
\hline
Storage & SSD for logs and database \\
\hline
OS & Linux (Ubuntu 22.04 LTS or Debian 12) \\
\hline
\end{longtable}

\textbf{Resource budget} (256 players):
\begin{itemize}
  \item CPU: game simulation $\sim$2 cores, voice $\sim$1 core, network $\sim$0.5 cores
  \item RAM: game state $\sim$500MB, voice buffers $\sim$200MB, total $\sim$2--3GB
  \item Network: $\sim$15--25Mbps sustained, 50--100Mbps burst
\end{itemize}

%---------------------------------------------------------------------
\section{Application Lifecycle}

\subsection{Client Lifecycle}

\begin{enumerate}
  \item \textbf{Initialization}:
    \begin{itemize}
      \item Parse command-line arguments
      \item Load configuration (settings, keybindings)
      \item Initialize raylib (window, audio, input)
      \item Initialize audio subsystem (microphone, speakers)
      \item Load assets (textures, fonts, sounds)
      \item Initialize database (local settings, cached data)
      \item Initialize network subsystem (ENet)
    \end{itemize}
  \item \textbf{Runtime}:
    \begin{itemize}
      \item Main loop with state machine dispatch
      \item Handle state transitions (menu $\rightarrow$ game $\rightarrow$ settings)
      \item Process input, update state, render
      \item Network polling and event handling
      \item Resource management (load/unload on state change)
    \end{itemize}
  \item \textbf{Shutdown}:
    \begin{itemize}
      \item Save user settings and preferences
      \item Disconnect from server (if connected)
      \item Unload assets and free resources
      \item Close database connections
      \item Shutdown ENet
      \item Close window and cleanup raylib
    \end{itemize}
\end{enumerate}

\subsection{Server Lifecycle}

\begin{enumerate}
  \item \textbf{Initialization}:
    \begin{itemize}
      \item Parse command-line arguments and config
      \item Initialize logger
      \item Install signal handlers (SIGINT, SIGTERM)
      \item Initialize database (SQLite, apply migrations)
      \item Initialize ENet
      \item Register with master server (if applicable)
      \item Initialize world state and spawn bots
    \end{itemize}
  \item \textbf{Runtime}:
    \begin{itemize}
      \item Fixed-timestep tick loop (30 Hz)
      \item Process network events
      \item Update simulation
      \item Broadcast snapshots
      \item Heartbeat to master server (every 30s)
      \item Log stats periodically
    \end{itemize}
  \item \textbf{Shutdown}:
    \begin{itemize}
      \item Handle SIGINT/SIGTERM gracefully
      \item Disconnect all clients with reason
      \item Save final session stats to database
      \item Deregister from master server
      \item Cleanup ENet
      \item Close database
      \item Flush logs
    \end{itemize}
\end{enumerate}

\subsection{Signal Handling}

\begin{itemize}
  \item SIGINT (Ctrl+C): graceful shutdown
  \item SIGTERM: graceful shutdown
  \item SIGHUP: reload config (optional, future)
\end{itemize}

%---------------------------------------------------------------------
\section{Screen State Machine}

\subsection{Screen States}

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.20\textwidth}|>{\raggedright\arraybackslash}p{0.70\textwidth}|}
\hline
\textbf{State} & \textbf{Description} \\
\hline
BOOT & Initial loading screen, load core assets, transition to MAIN\_MENU \\
\hline
MAIN\_MENU & Title screen with Play, Settings, Help, Credits, Exit buttons \\
\hline
AUTH & Login form (username/password), Discord OAuth, "Play as Guest" \\
\hline
SERVER\_BROWSER & List available servers, sort/filter, direct connect, quick match \\
\hline
LOBBY & Pre-game lobby, player list, chat, ready check, countdown \\
\hline
GAME & Active gameplay, HUD overlay, pause menu (if offline) \\
\hline
RESULTS & Post-game stats, leaderboard, play again or back to menu \\
\hline
SETTINGS & Graphics, audio, controls, network settings \\
\hline
HELP & Game controls, mechanics, arena zones, tips and strategies \\
\hline
CREDITS & Credits, licenses, acknowledgments \\
\hline
\end{longtable}

\subsection{State Transitions}

\begin{itemize}
  \item Stack-based navigation (back button pops state)
  \item Resource management: load on enter, unload on exit
  \item Fade in/out transitions
  \item Loading screen for heavy transitions
  \item Preserve state data when needed (e.g., settings persist)
\end{itemize}

\subsection{Input Handling}

\begin{itemize}
  \item Each screen handles its own input
  \item Global hotkeys: ESC (pause/back), F11 (fullscreen)
  \item Input buffering during transitions
\end{itemize}

%---------------------------------------------------------------------
\section{Dynamic Spawn Zones}

\subsection{Spawn Density Logic}

Spawn zones scale with player count to optimize gameplay density:

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.25\textwidth}|>{\raggedright\arraybackslash}p{0.35\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|}
\hline
\textbf{Lobby Size} & \textbf{Spawn Area} & \textbf{Density} \\
\hline
Small (1--64 players) & Wide area, more exploration & Low \\
\hline
Medium (65--128 players) & Moderate density & Balanced \\
\hline
Large (129--256 players) & Tight spawn zones & High conflict \\
\hline
\end{longtable}

\subsection{Spawn Algorithm}

\begin{enumerate}
  \item Calculate target density based on current player count
  \item Generate candidate spawn points within zone
  \item Filter by minimum distance from other players (340 units)
  \item Select spawn point (random or weighted by distance from center)
  \item Smooth transitions as lobby size changes
\end{enumerate}

\subsection{Configuration}

\begin{itemize}
  \item Configurable density curves (players per square unit)
  \item Min/max spawn radius bounds
  \item Minimum distance between spawns
  \item Respect existing arena zones (center, mid, outer)
\end{itemize}

%---------------------------------------------------------------------
\section{Client-Side Prediction}

\subsection{Prediction Model}

\begin{enumerate}
  \item \textbf{Input Application}: Client applies input immediately, updates local player position/velocity, sends input to server with sequence number
  \item \textbf{Server Reconciliation}: Server processes input, updates authoritative state, sends snapshot with processed input sequence
  \item \textbf{Correction}: Client compares predicted state vs server state. If mismatch, snap to server state. If error < threshold, interpolate smoothly. If error > threshold, hard snap (teleport).
\end{enumerate}

\subsection{Configuration}

\begin{itemize}
  \item Prediction error threshold (snap vs interpolate)
  \item Interpolation speed for corrections
  \item Handle packet loss (server didn't receive input)
  \item Handle out-of-order snapshots
  \item Prevent prediction drift over time
\end{itemize}

%---------------------------------------------------------------------
\section{Engineering Review and Risk Assessment}

\subsection{Code Quality}

The implementation is compact (12 source files, approximately 1~400 lines of C), well-structured, and consistent. The separation of shared simulation from client and server is maintained without violation. No known memory leaks or undefined behavior; the code uses stack-allocated structs and fixed-size arrays throughout with no dynamic allocation beyond ENet internals.

\subsection{Identified Gaps and Risks}

\begin{longtable}{|>{\raggedright\arraybackslash}p{0.08\textwidth}|>{\raggedright\arraybackslash}p{0.30\textwidth}|>{\raggedright\arraybackslash}p{0.22\textwidth}|>{\raggedright\arraybackslash}p{0.28\textwidth}|}
\hline
\textbf{ID} & \textbf{Gap} & \textbf{Severity} & \textbf{Mitigation} \\
\hline
G1 & Spores not replicated in snapshots; online players cannot collect spores & High & Issue \#24: add spore replication to snapshot protocol \\
\hline
G2 & Snapshot max 32 players vs simulation max 256; invisible players in full matches & High & Issue \#25: expand snapshot capacity or add interest management \\
\hline
G3 & No client-side prediction, interpolation, or reconciliation & Medium & Issue \#28: add basic client-side prediction \\
\hline
G4 & Player name sent in Hello but not stored or displayed anywhere & Medium & Issue \#25: store names server-side, include in snapshots \\
\hline
G5 & No menu, server browser, or connection UX & Medium & Issues \#34, \#35, \#37: add screen state machine and lobby browser \\
\hline
G6 & No audio, sprites, or themed content & Low & Future: art pass after core features complete \\
\hline
G7 & Deterministic PRNG seeded with constant & Low & Issue \#36: seed from time or allow config override \\
\hline
G8 & No RTT display on client despite ping/pong support & Low & Issue \#29: wire ping RTT to HUD status line \\
\hline
G9 & No CI, automated testing, or linting & Medium & Resolved: CI pipeline with clang-format, cppcheck, builds \\
\hline
G10 & No authentication system & High & Issue \#25: implement username/password + Discord OAuth + anonymous \\
\hline
G11 & No voice or text chat & Medium & Issues \#31, \#32: implement text and voice chat \\
\hline
G12 & No application lifecycle management & Medium & Issue \#36: implement structured lifecycle \\
\hline
\end{longtable}

%---------------------------------------------------------------------
\section{Engineering Roadmap}
\label{sec:roadmap}

The roadmap is organized by priority and dependency order. Each milestone corresponds to one or more GitHub issues.

\subsection{Phase 1: Core Infrastructure}

\begin{itemize}
  \item \textbf{Issue \#24}: Connection and session management (disconnect-and-rejoin model)
  \item \textbf{Issue \#25}: Authentication system (username/password + Discord OAuth + anonymous)
  \item \textbf{Issue \#27}: Message channel architecture (5 channels with reliability tiers)
  \item \textbf{Issue \#36}: Application lifecycle management (startup, runtime, shutdown)
  \item \textbf{Issue \#37}: Screen/page state machine (9 screens with transitions)
  \item \textbf{Issue \#38}: Game state management and persistence
\end{itemize}

\subsection{Phase 2: Networking and Performance}

\begin{itemize}
  \item \textbf{Issue \#26}: Dynamic spawn zones based on lobby size (256 players)
  \item \textbf{Issue \#28}: Basic client-side prediction (apply input immediately, reconcile)
  \item \textbf{Issue \#29}: Latency targets and monitoring (<100ms goal, 150ms warning)
  \item \textbf{Issue \#30}: Throughput optimization (45--50 msgs/sec per client, 12,800+ total)
  \item \textbf{Issue \#33}: Server hardware requirements (4 cores, 8GB RAM)
\end{itemize}

\subsection{Phase 3: Communication Features}

\begin{itemize}
  \item \textbf{Issue \#31}: Live text chat (global chat, reliable delivery)
  \item \textbf{Issue \#32}: Live voice chat (Opus codec, push-to-talk, custom implementation)
\end{itemize}

\subsection{Phase 4: Server Discovery and Matchmaking}

\begin{itemize}
  \item \textbf{Issue \#34}: Lobby browser and server list (master server)
  \item \textbf{Issue \#35}: Basic matchmaking (auto-assign to least-loaded/lowest-ping server)
\end{itemize}

\subsection{Phase 5: Online Spore Economy (G1)}

Add spore state to server snapshots so online matches include the full growth loop.

\begin{itemize}
  \item Extend \texttt{ShroomSnapshotPacket} or add a new \texttt{ShroomWorldStatePacket} type that includes spore data.
  \item Update \texttt{ApplyNetworkSnapshot} to populate \texttt{world.spores} from network data.
  \item Ensure spore collection is resolved authoritatively on the server and the client applies server spore state.
  \item Files affected: \texttt{protocol.h}, \texttt{server/main.c}, \texttt{client/net.c}, \texttt{client/game.c}.
\end{itemize}

\subsection{Phase 6: Snapshot Scaling (G2)}

Increase visible player capacity beyond 32.

\begin{itemize}
  \item Either: raise \texttt{SHROOM\_MAX\_SNAPSHOT\_PLAYERS} to 256 (packet size $\approx$ 10 KiB, acceptable over LAN/DC).
  \item Or: implement distance-based interest management to send only the nearest N players.
  \item Files affected: \texttt{protocol.h}, \texttt{server/main.c}, \texttt{client/net.c}.
\end{itemize}

\subsection{Phase 7: Polish and Content}

\begin{itemize}
  \item Seed PRNG from \texttt{time(NULL)} or add a config option.
  \item Display RTT in client HUD status line.
  \item Art direction pass: simple sprite rendering, audio stubs.
\end{itemize}

%---------------------------------------------------------------------
\section{Acceptance Criteria}

The v1 product is considered successful when:

\begin{itemize}
  \item a player can launch the client and navigate through the screen state machine (menu, settings, game),
  \item the player can authenticate via username/password, Discord OAuth, or play anonymously,
  \item the player can browse available servers and join a lobby (up to 256 players),
  \item the player can move immediately and grow via spore collection,
  \item larger entities can consume smaller ones under deterministic rules,
  \item the player respawns quickly after defeat,
  \item a headless server can host the simulation on UDP port 7777 with 256 concurrent players,
  \item a client can connect, complete handshake, submit input, and render server snapshots,
  \item text chat works for in-game communication,
  \item voice chat works with push-to-talk (Opus codec),
  \item latency is monitored and displayed (<100ms goal, 150ms warning),
  \item basic matchmaking auto-assigns players to optimal servers,
  \item the application lifecycle is structured (clean startup, runtime, shutdown),
  \item this specification document can be regenerated from the repository using \texttt{latexmk}.
\end{itemize}

\noindent Each issue in the roadmap extends this baseline with additional, verifiable criteria documented in the corresponding GitHub issue.

%---------------------------------------------------------------------
\section{Non-Goals for the Current Phase}

\begin{itemize}
  \item Deep progression systems (XP, levels, unlocks).
  \item Monetization design or microtransactions.
  \item Multi-region server deployment (single region for v1).
  \item High-fidelity art requirements before the core loop is proven.
  \item Large-scale persistent world simulation.
  \item Team-based game modes (free-for-all only for v1).
  \item Replay system or match history browser.
  \item Mobile client support.
\end{itemize}

%---------------------------------------------------------------------
\section{Document Maintenance}

This specification is the authoritative reference for the shroomio project. It must be updated whenever:

\begin{itemize}
  \item a tuning constant changes in \texttt{config.h} or \texttt{protocol.h},
  \item a new packet type or channel is introduced,
  \item the data model is extended,
  \item a roadmap milestone is completed,
  \item a new architectural module is added.
\end{itemize}

\noindent The source file lives at \texttt{design/shroomio-specification.tex} and is compiled with \texttt{make spec} (outputs to \texttt{dist/latex/}).

\end{document}
